Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-02 paper/ms.tex
update ms

%
% ms specific numbers
%

@@

%%%%%%%%%%%
% PLANETS %
%%%%%%%%%%%
\newcommand{\nkoiswithprot}{{$\sim$2{,}000}}
% number of KOIs in Q1-Q17 DR25 table that are not known false positives that have rotation periods
\newcommand{\nkoisnofpwithprot}{{$\sim$1{,}000}}
% fraction of KOIs that are not FPs with B+20 parameters
\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\approx$92\%}}

%%%%%%%%
% AGES %
%%%%%%%%
\newcommand{\nkepstarswithages}{NNN}
\newcommand{\nkoisnofpwithages}{NNN}
\newcommand{\nkoisgoldwithages}{NNN}

@@

%\title{Gyrochrone and Lithium Ages for Stars and Planets Observed by Kepler}
%Ages of the Kepler Stars and Planets
%\title{A Dearth of Young Stars and Planets in the Kepler Field}
\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}

@@

\author{coauthors!}

% <250 words
\begin{abstract}
  Recent analyses of FGK stars in open clusters have revealed how
  stellar rotation rates and lithium content evolve over the first few
  billion years.
  %
  Here we apply this knowledge to stars observed by Kepler to compute
  ages using rotation and lithium.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid; these rotation-based ages are
  $\approx$15-20$\times$ more precise than isochrone ages in our
  sample of FGK stars.
  %
  The independent lithium age estimates agree with the rotation-based
  ages for $\approx$90\% of stars.  However, the timescales of lithium
  depletion are such that rotation-based ages are more precise for
  most of our stellar sample.
  %
  Our resulting catalog of planetary ages includes
  \nplyounggyrotwosigma\ planets younger than 1\,Gyr at 2$\sigma$
  (\nplyounggyrothreesigma\ planets at 3$\sigma$).
  %
  More broadly, we find that the age distribution of stars in the
  Kepler field decreases linearly from 3\,Gyr ago to present day,
  for both planet hosts as well as for the overall stellar sample.
  %
  While this stellar age distribution has implications for the
  star formation history of our Galaxy, it also bounds the relative
  balance of young and old planets detectable through
  transit surveys of nearby stars in the thin disk.
  %
\end{abstract}

@@

%      and \citet{Petigura_2022} have reported trends between small-planet
%      occurrence and both stellar metallicity and mass; these parameters
%      seem, at first glance, to be more important predictors of bulk changes
%      in the planetary population than age.  Given these known correlations,
%      it can be useful to consider population models in which variables such
%      as metallicity, mass, and age are added in succession
%      \citep[e.g.][]{Thorngren_2021}.  An alternative, and perhaps more
%      rigorous approach, is to jointly infer the number of planets per star
%      as a function of both planetary and stellar parameters, the latter
%      including age \citep[e.g.][]{2023AJ....166..209M}.

%     While the youngest 0.1\% of stars ($\lesssim$10\,million
%     years; Myr) are perhaps the most constraining for studies of planet
%     formation, they are extremely challenging targets for close-in
%     exoplanet detection 
%     \citep[see e.g.][]{Damasso_2020,Bouma_2020_ptfo,Donati_2020}.  Studies of
%     the early orbital, structural, and atmospheric evolution of exoplanets
%     are perhaps most technically feasible for the 
%     $\approx$1\% of stars
%     with ages between 10 and 100\,Myr, or perhaps even the 10\% of stars
%     younger than 1\,Gyr.  However, since stellar ages are roughly
%     uniformly distributed in the solar neighborhood, there is a limit on
%     the sample size of young planets that will be detectable in these age
%     regimes, particularly relative to their older breathren.

%      Theoretical predictions for planet evolution in the Kepler sample are
%      most important in the context of mini-Neptunes and super-Earths, since
%      these are the most abundant planets.  Pertinent predictions include:
%      {\it a)} that mini-Neptunes should cool and contract, roughly
%      following a $R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}; {\it
%      b)} that large and close-in mini-Neptunes at the edge of the
%      photoevaporation desert should shrink over the atmospheric loss
%      timescale, $t_{\rm loss}$ \citep{Owen_Lai_2018}; {\it c)} that the
%      abundance of 4-8\,R$_\oplus$ planets with masses of 5-30\,M$_\oplus$
%      should decrease over $t_{\rm loss}$---the upper-envelope of the
%      mini-Neptune size distribution should corresponding decrease,
%      e.g.~from $\approx$4.5\,R$_\oplus$ at $\approx$200\,Myr to
%      $\approx$3.5\,R$_\oplus$ at $\approx$700\,Myr
%      \citep[e.g.][]{Rogers_2021}; {\it d)} that the super-Earth to
%      mini-Neptune ratio, $\theta$, should increase over $t_{\rm loss}$
%      \citep[e.g.][]{Rogers_2021}, with the mini-Neptune abundance
%      decreasing by at least a factor of two or three, and the super-Earth
%      abundance increasing in tandem; {\it e)} that the
%      upper boundary of the super-Earth population should increase over the
%      same timescale as $\theta$ is increasing, since more massive
%      super-Earth cores should be able to retain their envelopes for longer
%      than less massive cores.
%      
%      The details of the predictions are flexible, in that they depend on
%      the distributions of planetary core masses, initial atmospheric mass
%      fractions, and initial entropies of the atmospheres after disk
%      dispersal.  For instance, different sources of heating---from the star
%      \citep{Owen_Wu_2013,Lopez_Fortney_2014,Jin_2014}, planetary interior
%      \citep{Gupta_2019}, or even giant impacts
%      \citep{Biersteker_Schlichting_2019}---can produce very different
%      histories for the size evolution of planets at these early times
%      \citep[e.g.][]{Owen_2020}.  Different internal compositions also alter
%      the predicted size evolution \citep[e.g.][]{2024NatAs.tmp...33B}.  The
%      generic prediction that the ``upper envelope'' of the mini-Neptune
%      size distribution should shrink due to Kelvin-Helmholtz cooling is
%      hard to avoid; predictions however concerning the ``radius valley''
%      and movement of planets across it are dependent on the timescales and
%      prevalence of atmospheric mass loss.  In specific scenarios for
%      initial atmospheric accretion \citep{Lee_2022} and early atmospheric
%      evolution \citep{2024MNRAS.529.2716R} the radius gap might form
%      without needing to invoke post-disk dispersal mass loss at all.  In
%      such scenarios, there could be very little evolution of the planetary
%      size distribution between 1 and 4\,\rearth\ over the first gigayear.
%      
%      Can stellar age measurements provide evidence for or against
%      these different models?  Neither {\it (a)} nor {\it (b)} has yet had a
%      convincing---or even suggestive--- detection reported (see
%      \citealt{Petigura_2022}).  While an apparent over-abundance of large
%      very young planets ({\it (c)}) has been noted by multiple authors
%      \citep[e.g.][]{2017AJ....153...64M,Bouma_2020}, its presence is almost certainly
%      exacerbated by a selection effect against sub-Neptune sized planet
%      detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022b}.  Recent studies have
%      nonetheless found that when accounting for completeness, there is
%      $\approx$3$\sigma$ evidence that hot mini-Neptunes are more prevalent in
%      $\approx$100-500\,Myr old K2 \citep{2023AJ....166..248C} and TESS
%      \citep{2024arXiv240303261V} stars than in the older Kepler sample
%      \citep[e.g.][]{2020AJ....159..248K}.
%      %TODO: more kepler citations...
%      
%      Regarding the observed mini-Neptune to super-Earth ratio, {\it (d)} and {\it
%      (e)} were explored by \citet{Berger_2020b_rpage} and the twin
%      \citet{David_2021} and \citet{Sandoval_2021} studies.  Both
%      \citet{Berger_2020b_rpage} and \citet{Sandoval_2021} reported that $\theta$
%      increases with time, at a statistical significance of $\approx$2-3$\sigma$.
%      For \citet{Berger_2020b_rpage}, this conclusion was reached by comparing the
%      sizes of 85 planets with median isochronal ages younger than 1\,Gyr with a
%      property-matched sample of planets older than 1\,Gyr.  In
%      \citet{Sandoval_2021}, the conclusion was reached by computing the ratio while
%      performing a Monte Carlo re-sampling procedure from the isochronal ages and
%      planetary sizes computed by \citet{Fulton_Petigura_2018_cks_vii}.  Finally,
%      effect {\it (e)} was reported by \citet{David_2021}, who found that the average
%      location of the radius valley shifted from $\approx$1.8\,R$_\oplus$ to
%      $\approx$2.0\,R$_\oplus$ between 1.8 and 3.2\,Gyr.
%      
%      The fact that stellar ages---and especially isochronal ages---are
%      correlated with stellar masses and metallicities is a source of stress
%      for many in this subfield.  \citet{Sandoval_2021} for instance noted a
%      strong trend in $\theta$ as a function of stellar metallicity in
%      addition to the trend they noted in age, and cautioned that this could
%      also explain their observations.  Independently, \citet{Petigura_2018}
%      and \citet{Petigura_2022} have reported trends between small-planet
%      occurrence and both stellar metallicity and mass; these parameters
%      seem, at first glance, to be more important predictors of bulk changes
%      in the planetary population than age.  Given these known correlations,
%      it can be useful to consider population models in which variables such
%      as metallicity, mass, and age are added in succession
%      \citep[e.g.][]{Thorngren_2021}.  An alternative, and perhaps more
%      rigorous approach, is to jointly infer the number of planets per star
%      as a function of both planetary and stellar parameters, the latter
%      including age \citep[e.g.][]{2023AJ....166..209M}.  Such analyses are
%      most effective when the measurement precision of these different
%      parameters is as high as possible.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-4\,Gyr, and can also provide useful age limits at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock is quite old
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  
In the context of Kepler,
empirical models have been previously applied to derive rotation-based
ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015}.
There have been major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)}]
  \item The depth of the Kepler Object of Interest catalog, as well as our understanding of which of these objects are
    false positives;
  %
  \item Our empirical understanding of how rotation periods of stars
    in open clusters evolve
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation periods as a function of time
    \citep{Bouma_2023}.  
    The new model is broadly consistent with older calibrations
    \citep[e.g.][]{Mamajek_2008} for Sun-like stars.  However, for
    late G dwarfs and K dwarfs, this age-calibration differs by
    factors of a few from previous estimates;
  %
  \item Our ability to identify binary stars, which can potentially
    problematic for estimates of rotation-based ages.
\end{enumerate}
Today, rotation-based ages for FGK stars below $\approx$4\,Gyr are
empirically tied to the cluster age scale and rotation sequence, and
are limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

@@

\begin{figure*}[!t]
  \begin{center}
    \subfloat{
      \includegraphics[width=0.49\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
    }
% 
%   \subfloat{
%     \includegraphics[width=0.4\textwidth]{st_params_M_G_vs_adopted_Teff.pdf}
%   }
%   \vspace{-0.5cm}
    \subfloat{
    \includegraphics[width=0.49\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
    }
%   \subfloat{
%     \includegraphics[width=0.5\textwidth]{st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf}
%   }
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf The stars.}  Our analysis focuses on stars observed
    by Kepler with previously reported rotation periods (dark gray
    points).  About half of these rotators are suitable for
    gyrochronology (black points), based on factors including their
    temperatures, surface gravities, (non-)binarity, and proximity to
    the main sequence.  The full list of criteria is given in
    Section~\ref{subsec:flags}.  A few percent of the remainder are
    either ``confirmed'' or ``candidate'' Kepler Objects of Interest
    that meet an additional set of planetary quality criteria (blue points;
    Section~\ref{subsec:plflags}).  
% We independently derive
%    lithium-based ages for a largely-overlapping setting of KOIs based
%    on high-resolution spectroscopy.
    Surface gravities and effective
    temperatures are primarily from
    \citet{Berger_2020a_catalog}.
  }
  \label{fig:stellarprops}
\end{figure*}

@@

\begin{figure*}[!t]
  \begin{center}
    \subfloat{
      \includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
    }
    % 
    %   \vspace{-0.5cm}
    \subfloat{
      \includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
    }
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  The periods are from \citet{Santos_2019}
    and \citet{Santos_2021} (\citetalias{Santos_2019} and
    \citetalias{Santos_2021}).  The gray lines are ``mean fits'' from
    Figure~\ref{fig:models} to the slow rotation sequences of M67,
    Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open
    clusters.  The stellar sample shows only apparently single stars
    near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
    3800$<$$T_{\rm eff}$/K$<$6200, that are not flagged as eclipsing
    binaries.  The planet sample has the same stellar cuts, and shows
    confirmed and candidate planets from the cumulative KOI table,
    requiring them to have a MES$>$10.
  }
  \label{fig:prot_vs_teff}
\end{figure*}

\begin{figure}[!t]
  \begin{center}
    \leavevmode
    \includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
      for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected between 2009 and 2024.  Lines are the
    ``mean'' isochrones from the same models as in
    Figure~\ref{fig:models}; the intrinsic dispersion beyond
    $\approx$1\,Gyr becomes much larger than changes in the mean.  The
    logarithmic scale highlights the dynamic range of the detections
    and upper limits.  Some stars with lithium detections do not have
    detected rotation periods, and vice-versa.
    \label{fig:li_vs_teff}
  }
\end{figure}

@@

%Crossmatching the Q1-Q17 DR25 KOI table against our stars with
%rotation periods yields \nkoiswithprot\ KOIs with rotation periods, of
%which \nkoisnofpwithprot\ are not known false postives.  These objects
%will comprise the core data for our gyrochronology analysis.

@@

\subsection{Tally}
\label{subsec:tally}

@@

\begin{figure*}[!t]
  \begin{center}
    \leavevmode
    \includegraphics[width=\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage3200.pdf}
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf Kepler's demographic cliff,} visible in the rotation-derived
    ages of stars (left) and planet hosts (right) in the Kepler field.
    The lightly colored histograms correspond to all
    \nuniqstarsantosrotteffcut\ $T_{\rm eff}\in[3800,6200]\,$K Kepler
    stars with reported rotation periods, and all
    \nplhoststarwgyroage\ planet-hosting stars with reported rotation
    periods.
    The darker histograms impose necessary quality cuts on factors
    such as binarity, crowding, and the star's evolutionary state
    (Section~\ref{subsec:flags}).
    In detail, these ``histograms'' are resampled from the posterior
    probability distributions.
    Completeness is near unity at t$\lesssim$3\,Gyr
    \citep{2022ApJ...937...94M}.
    {\bf todo: numbers dont match.  why?}
    \label{fig:hist_tgyro}
  }
\end{figure*}

@@

Figure~\ref{fig:hist_tgyro} shows the gyrochone age distributions of
stars and planet-hosts in the Kepler field.
We truncated the plots near $3$\,Gyr due to arguments that suggest the
detectability of rotation signals is near-unity up to this age for
Sun-like stars, and decreases for older ages
\citep{2022ApJ...937...94M}.
Both the stellar and planetary samples point toward a star formation
rate in the Kepler field that has slowed over the past few gigayears.
{\it LGB pending words: allude to findings on MW SFH from chemistry,
isochrones, and asteroseismology}.

@@

%\begin{figure*}[!t]
% \begin{center}
%   \subfloat{
%     \includegraphics[width=0.5\textwidth]{rp_vs_period_scatter_20240415_colorbyage_strictyoung.pdf}
%   }
%   \subfloat{
%     \includegraphics[width=0.5\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
%   }
% 
%   \subfloat{
%   \includegraphics[width=0.5\textwidth]{rp_vs_period_scatter_20240415_colorbyage_anyyoung.pdf}
%   }
%   \subfloat{
%     \includegraphics[width=0.5\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
%   }
% \end{center}
% \vspace{-0.5cm}
%
% \caption{
%   {\bf Radii, orbital periods, and ages of transiting exoplanets}.
%    Planets younger than 1\,Gyr are emphasized. 
%    Two possible selection functions are suggested...
%   \label{fig:rp_period_age_results}
% }
%\end{figure*}

@@

\begin{figure*}[!t]
  \begin{center}
    %   \subfloat{
    %     \includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
    %   }
    %   
    %   \subfloat{
    \includegraphics[width=0.8\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
    %   }
  \end{center}
  \vspace{-0.5cm}

@@

\begin{enumerate}
  \item Two-sided posteriors available for both $t_{\rm gyro}$ and
    $t_{\rm Li}$: if their values were consistent within
    2$\sigma$, we judged them to be consistent; if they were
    consistent within 2-3$\sigma$, we suggest them to be ``maybe''
    consistent.  Otherwise, they are in disagreement.
\item If $t_{\rm Li}$ is a lower limit, we compare the
  $t_{\rm Li}$ lower limit with the $1\sigma$ upper limit from $t_{\rm
    gyro}$; if the two overlapped, we judge the age estimates to be
    consistent.
\item If $t_{\rm Li}$ provides a two-sided posterior, and no rotation
  is reported, we judge the age estimates to be inconsistent.
  This is because this lithium constraint can only be provided for G and K
  stars $\lesssim$2\,Gyr old, and Kepler should have been sensitive to
  the rotation periods of such stars.
\end{enumerate}

@@

\subsubsection{Systems with Rotation and Lithium in Agreement}
%
Whenever both rotation and lithium find a star to be in a
narrowly-constrained age range, confidence in the true age greatly
increases, due to the independence of the two measurement techniques.
The empirical precision for the inferred ages as a function of $P_{\rm
rot}$, $EW_{\rm Li}$, and $T_{\rm eff}$ has been mapped out by
\citet{Bouma_2023} and \citet{Jeffries_2023} for rotation and lithium
respectively.  The region of greatest overlap is dictated by the
timescale for lithium depletion: for stars with temperatures between
4500-5500\,K with ages $\lesssim$500\,Myr, we can hope to obtain
independent two-sided constraints.

@@

\subsubsection{Need spectra}
Kepler-1938...
Kepler-762... (a HJ!)
Kepler-1320

\subsubsection{Other Notables}

Kepler~411, a $\approx$780$\pm$170\,Myr early K dwarf with three
transiting planets at 3.0, 7.8, and 58\,days, has a rare
architecture.  \citet{2019A&A...624A..15S} conducted a transiting
timing analysis of the system, confirming Kepler-411d and arguing for
the presence of a fourth non-transiting planet, Kepler-411e, based on
the TTVs seen in Kepler-411d and assuming that the system is nearly
coplanar.  Regarding the stellar age, \citet{2019A&A...624A..15S} used
the \citet{2007ApJ...669.1167B} relation and the measured rotation
period, which yielded an estimate of 212$\pm$31\,Myr.  However, the
star's temperature ($\approx$4900\,K) puts it in a regime where the
\citeauthor{2007ApJ...669.1167B} relation has been known to not fit
observations for quite some time
\citep[e.g.][Fig.~9]{2008ApJ...687.1264M}.  Our adopted age of
$\approx$780$\pm$170\,Myr agrees with the open cluster calibrators.

Kepler-63b is interesting in that it is {\it missing} from our
catalog.  This 6.1\,$R_\oplus$ ``sub-Saturn'' has been shown to be in
a polar orbit based on both the Rossiter-McLaughlin effect and a
starspot-crossing analysis \citep{2013ApJ...775...54S}.  With $T_{\rm
eff}=5576\pm50$\,K, and a rotation period of 5.40\,days, we calculate
a rotation-based age of $t_{\rm gyro}$=$251^{+80}_{-90}$\,Myr,
qualitatively consistent with the estimates from
\citet{2013ApJ...775...54S}, though with larger uncertainties because
our age accounts for the intrinsic scatter in the open cluster
calibration sequences \citep{Bouma_2023}.  
While the usual catalogs (\citetalias{Santos_2019},
\citetalias{Santos_2021}, \citetalias{McQuillan_2014}, and
\citetalias{Mazeh_2015}) do not report this star's rotation period.
\citetalias{Santos_2021} flagged the system as ``potentially'' having
rotation-based modulation.
{\bf todo: how many comparable missed cases are there?}.
%TODO FIXME it's bc you need to do mcquillan 2013....
%
%... David+2021 used:
% Angus+2018... McQuillan+2013... Mazeh+2015... Walkowicz&Basri2013...

\subsection{Evolutionary trends in time}

{\it Kelvin-Helmholtz cooling}
$R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}

{\it Early carving of the photoevaporation desert}
\citep{Owen_Lai_2018}

{\it Rapidly decreasing abundance of big puffy planets}
Kepler not enough?

{\it Mini-Neptunes turning into Super-Earths}
\citep[e.g.][]{Rogers_2021}

{\it Boundary of the Super-Earth population moving up in time}
\citep{David_2021}

@@

\subsection{``Galactic Archaeology'' Comparison}
Zasowski et al fitted a chemical-isochrone model to APOGEE, X, Y, Z
data and presented a model for the star formation history of the MW.
The gist is a spike at z=2, and a gradual decrease since then, by
about a few of ten.

@@

\subsection{Isochrone Age Comparison}

\subsection{Lithium Age Comparison}

\subsection{Gyrochronal Age Comparison}
Searching the literature for gyrochronal analyses of the Kepler field,
the most relevant studies seemed to be those of
\citet{Walkowicz_2013}, \citet{Reinhold_2015}, and 
\citet{David_2021}.

>>>>

@@

%%%%%%%%%%%
% PLANETS %
%%%%%%%%%%%
% fraction of KOIs that are not FPs with B+20 parameters
\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\approx$92\%}}

@@

%\title{Gyrochrone and Lithium Ages for Stars and Planets Observed by Kepler}
%\title{Ages of the Kepler Stars and Planets}
%\title{A Dearth of Young Stars and Planets in the Kepler Field}
\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}

@@

\author{coauthors, to include but not limited to
Hillenbrand, Howard, Isaacson, Palumbo}

% <250 words
\begin{abstract}
  %Recent analyses of FGK stars in open clusters have revealed how
  %stellar rotation rates and lithium content evolve over the first few
  %billion years.
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate, and its lithium
  content, can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium
  is measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid; these rotation-based ages are
  roughly an order of magnitude more precise than isochrone ages in
  this sample of FGK stars.
  %
  The independent lithium ages agree with the rotation-based ages for
  $\approx$90\% of cases, suggesting that most of the rotation-based
  ages are not only precise, they are also accurate.
  %However, the timescales of lithium depletion are such that
  %rotation-based ages are more precise for most of our stellar sample.
  %
  Our resulting catalog includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler
  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr over the
  past 3\,Gyr, for both planet hosts as well as for the overall
  stellar sample.
  %
  This ``demographic cliff'' has implications for the star formation
  history of our Galaxy, and it helps clarify the relative balance of
  young and old planets that have been detected in transit surveys of nearby
  stars in the Galaxy.
  %
\end{abstract}

@@

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-4\,Gyr, and can also provide useful age limits at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock is quite old
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  
In the context of Kepler,
empirical models have been previously applied to derive rotation-based
ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.
There have been major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)}]
  \item The extent of the Kepler Object of Interest catalog, as well
    as our understanding of which of these objects are false
    positives;
  %
  \item Our empirical understanding of how rotation periods of stars
    in open clusters evolve
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation periods as a function of time
    \citep{Bouma_2023}.  
    For Sun-like stars, the new model is broadly consistent with older
    calibrations \citep[e.g.][]{Mamajek_2008}.  However, for late G
    dwarfs and K dwarfs, the age-calibration differs by factors of a
    few;
  %
  \item Our ability to identify binary stars, which can potentially
    problematic for estimates of rotation-based ages.
\end{enumerate}
Today, rotation-based ages for FGK stars below $\approx$4\,Gyr are
empirically tied to the cluster age scale and rotation sequence, and
are limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

@@

\begin{figure*}[!t]
  \begin{center}
    \subfloat{
      \includegraphics[width=0.49\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
    }
    \subfloat{
    \includegraphics[width=0.49\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
    }
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf The stars.}  Our analysis focuses on stars observed by Kepler
    with previously reported rotation periods (dark gray points).
    About half of these rotators are suitable for gyrochronology
    (black points), based on factors including their temperatures,
    surface gravities, non-binarity, and proximity to the main
    sequence (see Section~\ref{subsec:flags}).   Some are
    ``confirmed'' or ``candidate'' Kepler Objects of Interest that
    meet an additional set of planetary quality criteria (blue points;
    Section~\ref{subsec:plflags}).  Surface gravities and effective
    temperatures are primarily from \citet{Berger_2020a_catalog}.
  }
  \label{fig:stellarprops}
\end{figure*}

@@

\begin{figure*}[!t]
  \begin{center}
    \subfloat{
      \includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
    }
    \subfloat{
      \includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
    }
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  The periods are from \citet{Santos_2019}
    and \citet{Santos_2021} (\citetalias{Santos_2019} and
    \citetalias{Santos_2021}).  The gray lines are ``mean fits'' from
    Figure~\ref{fig:models} to the slow rotation sequences of M67,
    Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open
    clusters.  The stellar sample includes only apparently single stars
    near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
    and 3800$<$$T_{\rm eff}$/K$<$6200.  The displayed planets require
    the same stellar cuts, and include only the confirmed and candidate planets
    described in Section~\ref{subsec:plflags}.
  }
  \label{fig:prot_vs_teff}
\end{figure*}

\begin{figure}[!t]
  \begin{center}
    \leavevmode
    \includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
      for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected between 2009 and 2024 comprising
    \nhireshours\,hours of open-shutter time.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}; the
    intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
    than changes in the mean.  The logarithmic scale highlights the
    dynamic range of the detections and upper limits.  Some stars with
    lithium detections do not have detected rotation periods, and
    vice-versa.
    \label{fig:li_vs_teff}
  }
\end{figure}

@@

@@

\subsection{Sample Summary}
\label{subsec:tally}

@@

\begin{figure*}[!t]
  \begin{center}
    \leavevmode
    \includegraphics[width=\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage3200.pdf}
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf Kepler's demographic cliff,} visible in the rotation-derived
    ages of stars (left) and planet hosts (right) in the Kepler field.
    The lightly colored histograms correspond to all
    \nuniqstarsantosrotteffcut\ $T_{\rm eff}\in[3800,6200]\,$K Kepler
    stars with reported rotation periods, and all
    \nplhoststarwgyroage\ planet-hosting stars with reported rotation
    periods.
    The darker histograms impose the necessary quality cuts on factors
    such as binarity, crowding, and the star's evolutionary state
    (Section~\ref{subsec:flags}).
    In detail, these ``histograms'' are resampled from the posterior
    probability distributions, and so have less resolution at
    $\lesssim$0.5\,Gyr, when the relative precision of rotation-based
    ages decreases.
    Completeness is near unity at t$\lesssim$3\,Gyr
    \citep{2022ApJ...937...94M}.
    {\bf todo: numbers dont match.  why?}
    \label{fig:hist_tgyro}
  }
\end{figure*}

@@

Figure~\ref{fig:hist_tgyro} shows the gyrochone age distributions of
stars and planet-hosts in the Kepler field.  We truncated the plots
near $3$\,Gyr due to arguments that suggest the detectability of
rotation signals is near-unity up to this age for Sun-like stars, and
decreases for older ages \citep{2022ApJ...937...94M}.  Both the
stellar and planetary samples point toward a star formation rate in
the Kepler field that has slowed over the past few gigayears.

@@

@@

\begin{figure*}[!t]
  \begin{center}
    \includegraphics[width=0.7\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
  \end{center}
  \vspace{-0.5cm}

@@

\begin{enumerate}
  \item Two-sided posteriors available for both $t_{\rm gyro}$ and
    $t_{\rm Li}$: if their values were consistent within 2$\sigma$, we
    judged them to be consistent; if they were consistent within
    2-3$\sigma$, we suggest them to be ``maybe'' consistent.
    Otherwise, they are in disagreement.
    %
  \item If $t_{\rm Li}$ is a lower limit, we compare the $t_{\rm Li}$
    lower limit with the $1\sigma$ upper limit from $t_{\rm gyro}$; if
    the two overlapped, we judge the age estimates to be consistent.
    %
  \item If $t_{\rm Li}$ provides a two-sided posterior, and no
    rotation is reported, we judge the age estimates to be
    inconsistent.  This is because this lithium constraint can only be
    provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
    have been sensitive to the rotation periods of such stars.
\end{enumerate}

@@

\subsubsection{New Young Systems}
%
Whenever both rotation and lithium find a star to be in a
narrowly-constrained age range, confidence in the true age greatly
increases, due to the independence of the two measurement techniques.
The empirical precision for the inferred ages as a function of $P_{\rm
rot}$, $EW_{\rm Li}$, and $T_{\rm eff}$ has been mapped out by
\citet{Bouma_2023} and \citet{Jeffries_2023} for rotation and lithium
respectively.  The region of greatest overlap is dictated by the
timescale for lithium depletion: for stars with temperatures between
4500-5500\,K with ages $\lesssim$500\,Myr, we can hope to obtain
independent two-sided constraints.

@@

%\subsubsection{Need spectra}
%Kepler-1938...
%Kepler-762... (a HJ!)
%Kepler-1320

@@

\subsection{The Thin Disk's Star Formation History}

Zasowski et al fitted a chemical-isochrone model to APOGEE, X, Y, Z
data and presented a model for the star formation history of the MW.
The gist is a spike at z=2, and a gradual decrease since then, by
about a few of ten.

@@

%\subsection{Isochrone Age Comparison}

%\subsection{Lithium Age Comparison}

\subsection{Caveats}

\subsection{Future Directions}
